```{r setup, include = FALSE}
opts_chunk$set(echo = FALSE
               , warning = FALSE
               , message = FALSE
               , fig.width = 16 / 2.54
               , fig.height = 6
 #              , dpi = 600 # use 600 for printing otherwise comment out
               , fig.cap = "")

tableStyle <- "rmarkdown" # for .docx use "multiline" otherwise "rmarkdown"
tableResults <- "asis"
excelPalette <- c("#4F81BD", "#C0504D", "#9BBB59", "#8064A2")
```

```{r function, include = FALSE}
source("R/01-function.r")
```

```{r source, include = FALSE, cache = TRUE}
source("R/02-load.r")
source("R/03-clean.r")
```

Rural Drink-Driving in [District]
=============================

All the crashes discussed in this document are fatal/serious-injury crashes.


### Numbers of crashes

[Input text here]

```{r pureCrashesTable, results = tableResults, cache = TRUE}
x <- dcast(crashes, urbanRural + alcohol ~ year, length, value.var = "count")
colnames(x)[1:2] <- c("Urban/Rural", "Alcohol a Factor")
pandoc.table(x
             , caption = "Number of crashes by whether alcohol was a factor"
             ,style = tableStyle
             , split.tables = Inf)
rm(x)
```

```{r pureCrashes, cache = TRUE}
plotLabels <- list("TRUE" = "alcohol a factor"
  , "FALSE" = "alcohol not a factor")

plotLabeller <- function(variable, value) {
  return(plotLabels[value])
}

ggplot(crashes, aes(year, group = urbanRural, fill = urbanRural)) + 
 geom_bar(binwidth = 1, position = "dodge", colour = "black") + 
  facet_grid(. ~ alcohol, labeller = plotLabeller) +
  ggtitle("Number of Crashes\nby whether alcohol was a factor") +
  ylab("number of crashes") +
  theme(legend.title = element_blank()) +
  scale_fill_manual(values = excelPalette)
rm(plotLabels)
rm(plotLabeller)
```


### Crash rate per 1000 km of road

Roads have been divided into four categories, according to the type of road they are, and the area they are in.

* urban roads (excluding highways)
* urban highways
* rural roads (excluding highways)
* rural highways

```{r roadTable, results = tableResults, cache = TRUE}
pandoc.table(SummaryT[c("urban", "rural"), c("road", "highway")]
             , caption = "Road in Tasman (km)"
             ,style = tableStyle
             , split.tables = Inf)
```

[Input text here]

```{r roadCrashRateTable, results = tableResults, cache = TRUE}
x <- dcast(crashes, urbanRuralRoadHighway + alcohol ~ year, sum, value.var = "countRoad")
colnames(x)[1:2] <- c("Urban/Rural\nRoad/StateHighway", "Alcohol a Factor")
pandoc.table(x
             , caption = "Crashes per 1000 km by type of road and whether alcohol was a factor"
             , round = 0
             , keep.trailing.zeros = TRUE
             , justify = c(rep("left", 2), rep("right", 5))
             ,style = tableStyle
             , split.tables = Inf)
rm(x)
```

```{r roadCrashRate, cache = TRUE}
ggplot(crashes, aes(year, weight = countRoad, group = urbanRuralRoadHighway
                    , fill = urbanRuralRoadHighway)) + 
  geom_bar(position = "dodge", colour = "black") + facet_wrap(.(alcohol)) +
  ggtitle("Crashes per 1000 km\nby type of road\nand whether alcohol was a factor") +
  ylab("crashes per 1000 km") +
  theme(legend.title = element_blank()) +
  scale_fill_manual(values = excelPalette)
```


### Crashes by population

Areas of [District] have been categorised into "urban" and "rural", based on a Statistics New Zealand dataset.  Demographic data are available giving the populations of urban and rural areas.  [Input text here]

```{r populationTable, results = tableResults, cache = TRUE}
pandoc.table(data.frame("Region" = c("urban", "rural")
                        , Population = SummaryT[c("urban", "rural")
                                                  , "population"])
             , caption = "Population of Tasman"
             , round = 0
             , keep.trailing.zeros = TRUE
             , justify = c("left", "right")
             , style = tableStyle
             , split.tables = Inf)
```


[Input text here]

```{r populationCrashRateTable, results = tableResults, cache = TRUE}
x <- dcast(crashes, urbanRural + alcohol ~ year, sum, value.var = "countPopulation")
colnames(x)[1:2] <- c("Urban/Rural\nRoad/StateHighway", "Alcohol a Factor")
pandoc.table(x
             , caption = "Crashes per 1000 people by urban/rural and whether alcohol was a factor"
             , round = 2
             , keep.trailing.zeros = TRUE
             , justify = c(rep("left", 2), rep("right", 5))
             , style = tableStyle
             , split.tables = Inf)
rm(x)
```

```{r populationCrashRate, cache = TRUE}
plotLabels <- list(
  "TRUE" = "alcohol a factor"
  , "FALSE" = "alcohol not a factor"
  )

plotLabeller <- function(variable, value) {
  return(plotLabels[value])
}

ggplot(crashes, aes(year, weight = countPopulation, group = urbanRural
                    , fill = urbanRural)) + 
  geom_bar(position = "dodge", colour = "black") +
  facet_grid(. ~ alcohol, labeller = plotLabeller) +
  ggtitle("Crashes per 1000 people\nby urban/rural and whether\nalcohol was a factor") +
  ylab("number of crashes per 1000 people") +
  theme(legend.title = element_blank()) +
  scale_fill_manual(values = excelPalette)
rm(plotLabels)
rm(plotLabeller)
```

Compare with the first graph, "Number of Crashes".  Weighting by population is a fairer estimation of the rural drink-drive problem.  A better weighting would take into account the number of licenced drivers and distance driven, however these data are not available.


### Time of day of crashes

[Input text here]

```{r crashesHourlyByNumberTable, results = tableResults, cache = TRUE}
x <- dcast(crashes[!is.na(crashes$hour) & crashes$alcohol == TRUE, ]
           ,  hour ~ urbanRural + alcohol, length, value.var = "count")
colnames(x) <- c("Hour of Day", "Rural", "Urban")
pandoc.table(x
             , caption = "Crashes where Alcohol was a Factor by hour of the day"
             , keep.trailing.zeros = TRUE
             , round = 0
             , justify = c("left", rep("right", 2))
             , style = tableStyle
             , split.tables = Inf)
rm(x)
```

```{r crashesHourlyByNumber, cache = TRUE}
ggplot(crashes[!is.na(crashes$hour) & crashes$alcohol == TRUE, ]
       , aes(hour, weight = count
             , fill = as.character(excelPalette[1]))) + 
  geom_bar(binwidth = 1, colour = "black") +
  scale_x_discrete(breaks = seq(0,24,4)) +
  facet_grid(urbanRural ~ .) +
  ggtitle("Crashes where Alcohol was a Factor\nby hour of the day") +
  ylab("number of crashes") +
  theme(legend.position="none") +
  scale_fill_manual(values = excelPalette)
```


[Input text here]

```{r crashesHourlyByPercentageTable, results = tableResults, cache = TRUE}
x <- dcast(crashes[!is.na(crashes$hour) & crashes$alcohol == TRUE, ]
           , hour ~ urbanRural, sum, value.var = "countCrashUrbanHour")
colnames(x) <- c("Hour of Day", "Rural", "Urban")
pandoc.table(x
             , caption = "% of Crashes where alcohol was a factor by hour of the day"
             , round = 0
             , justify = c("left", rep("right", 2))
             , keep.trailing.zeros = TRUE
             , style = tableStyle
             , split.tables = Inf)
```

```{r crashesHourlyCaption, include = FALSE, cache = TRUE}
crashesHourlyCaption <- "The bars indicate the percentage of all crashes that were drink-drive-related in that hour of the day, in that region, e.g. the right-most bar of the top graph shows that over 60% of rural crashes between 2300 hours and midnight were drink-driver related."
```

```{r crashesHourlyByPercentage, fig.cap = crashesHourlyCaption, cache = TRUE}
# crashes that are drink-drive-related, as a percentage of all crashes
ggplot(crashes[!is.na(crashes$hour) & crashes$alcohol == TRUE, ]
       , aes(hour, weight = countCrashUrbanHour
             , fill = as.character(excelPalette[1]), label = sum(countCrashUrbanHour))) + 
  geom_bar(binwidth = 1, colour = "black") +
  scale_x_discrete(breaks = seq(0,24,4)) +
  facet_grid(urbanRural ~ .) +
  ggtitle("% of Crashes where Alcohol was a Factor\nby hour of the day") +
  ylab("% of crashes") +
  theme(legend.position="none") +
  theme(panel.margin = unit(0, "pt")) +
  scale_fill_manual(values = excelPalette)
```


[Input text here]

```{r crashesWeeklyTable, results = tableResults, cache = TRUE}
x <- dcast(crashes[!is.na(crashes$hour) & crashes$alcohol == TRUE, ]
           , urbanRural + weekday + hour ~ ., sum, value.var = "countCrashUrbanWeekdayHour3")
# group into threes
x$hour3 <- cut(as.integer(x$hour), breaks = seq(0, 24, 3), right = FALSE, labels = c(paste(seq(0, 21, 3), "-", seq(3, 24, 3), sep = "")))
x <- ddply(x, .(urbanRural, weekday, hour3), function(x) (sum(x$`NA`)))
x <- dcast(x, urbanRural + weekday ~ hour3, sum, value.var = "V1")
colnames(x)[1:2] <- c("Urban/Rural", "Weekday")
pandoc.table(x
             , caption = "% of Crashes where Alcohol was a factor by weekday hour of the day"
             , round = 0
             , justify = c(rep("left", 2), rep("right", 8))
             , keep.trailing.zeros = TRUE
             , style = tableStyle
             , split.tables = Inf)
```

```{r crashesWeekly, cache = TRUE}
# make a copy of the crashes
x <- crashes
# centre on Thursday for plotting
x$weekday <- factor(x$weekday, levels = c("Wednesday"
                                          , "Thursday", "Friday"
                                          , "Saturday", "Sunday"
                                          , "Monday", "Tuesday"))
ggplot(x[!is.na(x$hour) 
         & x$alcohol == TRUE, ]
       , aes(hour, weight = countCrashUrbanWeekdayHour3
             , fill = as.character(excelPalette[1]))) + 
  geom_histogram(binwidth = 3, colour = "black") +
  scale_x_continuous(breaks = seq(0, 24 ,6), expand = c(0, 0), limits = c(0, 24)) +
  scale_y_continuous(expand = c(0, 0)) +
  facet_grid(urbanRural ~ weekday) +
  ggtitle("% of Crashes where Alcohol was a Factor\nby weekday and hour of the day") +
  xlab("hour of the day (in three-hour groups)") +
  ylab("% of crashes") +
  theme(legend.position="none") +
  theme(panel.margin = unit(0, "pt")
        , panel.border = element_rect(colour = "black", fill = "transparent")) +
  scale_fill_manual(values = excelPalette)
# tidy
rm(x)
```


### Age / ethnicity of drivers-at-fault

[Input text here]

```{r ageTable, results = tableResults, cache = TRUE}
x <- dcast(drivers[!is.na(drivers$age) & drivers$fault == TRUE & drivers$alcohol == TRUE, ]
           , urbanRural + age ~ ., sum, value.var = "countAgeGroupPopulation")
x <- x[x$age >= 15, ]
# group into fives
x$age5 <- cut(x$age, breaks = seq(15, 79, 5), right = FALSE, labels = c(paste(seq(15, 69, 5), "-", seq(19, 69, 5), sep = ""), "70+"))
x$age5[is.na(x$age5)] <- "70+"
x <- ddply(x, .(urbanRural, age5), function(x) (sum(x$`NA`)))
x <- dcast(x, age5 ~ urbanRural, sum, value.var = "V1")
colnames(x)[1:3] <- c("Age Group", "Rural", "Urban")
pandoc.table(x
             , caption = "Drink-Drivers in Crashes per 1000 people by age in 5-year bands"
             , round = 2
             , justify = c("left", rep("right", 2))
             , keep.trailing.zeros = TRUE
             , style = tableStyle
             , split.tables = Inf)
rm(x)
```

```{r age, cache = TRUE}
x <- drivers[!is.na(drivers$age) & drivers$fault == TRUE & drivers$alcohol == TRUE, ]
ggplot(x, aes(age, weight = countAgeGroupPopulation
              , fill = as.character(excelPalette[1]))) +
  geom_histogram(binwidth = 5, colour = "black") +
  scale_x_continuous(breaks = seq(20, 80, 10), limits = c(15, max(x$age))) +
  facet_grid(urbanRural ~ .) +
  ggtitle("Drink-Drivers in Crashes per 1000 people\nby age in 5-year bands") +
  xlab("age in 5-year bands") +
  ylab("number of drink-drivers in crashes per 1000 people") +
  theme(legend.position="none") +
  scale_fill_manual(values = excelPalette)
# tidy
rm(x)
```



[Input text here]

```{r ethnicityTableDrivers, results = tableResults, cache = TRUE, }
x <- ddply(drivers[drivers$fault == TRUE & drivers$alcohol == TRUE, ]
           , .(ethnicity, urbanRural), function(x) (sum(x$count)))
x <- dcast(x, ethnicity ~ urbanRural, sum, margins = TRUE)
# nicer Total row and column headings
colnames(x) <- c("Ethnicity", "Rural Drivers", "Urban Drivers", "Total")
x$Ethnicity <- as.character(x$Ethnicity)
x$Ethnicity[nrow(x)] <- "Total"
# remove row names because pander can't ignore them
rownames(x) <- NULL
# print
pandoc.table(x, caption = "Drink-Drivers At Fault by Ethnicity"
             , emphasize.strong.rows = 5, emphasize.strong.cols = 4
             , round = 0
             , justify = c("left", rep("right", 3))
             , keep.trailing.zeros = TRUE
             , style = tableStyle
             , split.tables = Inf)
# tidy
rm(x)
```

```{r ethnicityTablePopulation, results = tableResults, cache = TRUE}
x <- dcast(mDataT[mDataT$variable %in% c("European", "NZ Maori", "Other"
                                        , "Pacific Islander"), ]
      , variable ~ urbanRural, sum, na.rm = TRUE, margins = TRUE)
# remove NA column
x <- x[, -4]
# order alphabetically leaving (all) at the bottom
x$variable <- as.character(x$variable)
x <- x[c(order(x$variable)[2:5], 5), ]
# nicer Total row and column headings
x$variable[length(x$variable)] <- "Total"
colnames(x) <- c("Ethnicity", "Rural Population", "Urban Population", "Total")
# remove row names because pander can't ignore them
rownames(x) <- NULL
# print
pandoc.table(x, caption = "Population by Ethnicity"
             , emphasize.strong.rows = 5, emphasize.strong.cols = 4
             , round = 0
             , justify = c("left", rep("right", 3))
             , keep.trailing.zeros = TRUE
             , style = tableStyle
             , split.tables = Inf)
# tidy
rm(x)
```

```{r ethnicityTable, results = tableResults, cache = TRUE}
x <- dcast(drivers[!is.na(drivers$age) & drivers$fault == TRUE & drivers$alcohol == TRUE, ]
           , urbanRural + ethnicity ~ ., sum, value.var = "countUrbanEthnicity")
colnames(x) <- c("Urban/Rural", "Ethnicity", "Drink-Drivers per 1000")
pandoc.table(x
             , caption = "Drink-drivers in crashes per 1000 people by urban/rural and ethnicity"
             , round = 2
             , justify = c(rep("left", 2), "right")
             , keep.trailing.zeros = TRUE
             , style = tableStyle
             , split.tables = Inf)
# tidy
rm(x)
```

```{r ethnicity, cache = TRUE}
x <- drivers[!is.na(drivers$age) & drivers$fault == TRUE & drivers$alcohol == TRUE, ]
ggplot(x, aes(ethnicity, weight = countUrbanEthnicity
              , fill = ethnicity)) +
  geom_bar(colour = "black") + 
  facet_grid(. ~ urbanRural) +
  ggtitle("Drink-Drivers in Crashes per 1000 people\nby urban/rural and ethnicity") +
  xlab("ethnicity") +
  ylab("number of drink-drivers in crashes per 1000 people") +
  theme(legend.position="none") +
  scale_fill_manual(values = excelPalette)
# tidy
rm(x)
```


### Discussion

[Input text here]